---
title: "Assignment 7 Report"
author: "Ayesha Zafar"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
---

# 7. Gaussian processes

---

### **7.1 Build a Stan implementation for the following model y(x)∼N(f(x),σ2)f(x)∼GP(0,KSE(α,λ))σ∼Γ(2,1),where KSE(α,λ) is the squared exponential covariance kernel with variance α2 and length-scale λ. Set the natural logarithm of tree circumference as y, and age as x. Set suitable values for the hyperparameters. Fit the model on the data.**

The code below implements a Gaussian Process model with a squared exponential covariance kernel in stan. 
The gaussian process is used to model the natural logarithm of tree circumferences (y) as a function of tree age (x). The squared exponential kernel defines the covariance structure with parameters α (variance) and λ (length scale), while σ accounts for observation noise.

The stan model specifies following things:

1. Kernel Function: The squared exponential kernel to capture smooth variations in y(x).

2. Latent Function f(x): Multivariate normal prior with covariance matrix from the kernel.

3. Observation Model: y ∼ N(f,σ^2), with added noise.

4. Priors: For α, λ, and σ, to ensure regularization.

Lastly, the model is fit to the data, estimating posterior distributions for the kernel parameters and noise. It also tells the latent function values f(x), which describes the underlying trend in the data.

```{r}
library(rstan)

# Loading data from data7.txt
data <- read.csv("data7.txt", stringsAsFactors = FALSE)

# Checking columns are correctly typed
data$circumference <- as.numeric(data$circumference)
data$age <- as.numeric(data$age)
summary(data)

# Preparing variables x and y
x <- data$age
y <- log(data$circumference)

# Check for NA values
if (any(is.na(y))) {
  stop("Log of circumference contains NA values.")
}

# Setting stan model
stan_code <- "
data {
  // No of data points
  int<lower=1> N;               
  // Tree age
  vector[N] x;                  
  // Log circumference
  vector[N] y;                  
}
parameters {
  // Variance
  real<lower=0> alpha;          
  // Length-scale
  real<lower=0> lambda;         
  // Noise
  real<lower=0> sigma;          
  // Latent function values
  vector[N] f;                  
}
model {
  // α, λ, and σ Priors
  alpha ~ normal(0, 5);
  lambda ~ normal(0, 5);
  sigma ~ gamma(2, 1);
  
  // Setting and filling squared exponential covariance matrix
  matrix[N, N] K;
  for (i in 1:N) {
    for (j in 1:N) {
      K[i, j] = alpha^2 * exp(-0.5 * square(x[i] - x[j]) / lambda^2);
    }
  }
  K += diag_matrix(rep_vector(sigma^2, N)); // Add noise variance
  
  // Likelihood
  f ~ multi_normal(rep_vector(0, N), K);
  y ~ normal(f, sigma);
}
generated quantities {
  // Posterior predictive
  vector[N] y_pred;             
  for (i in 1:N) {
    y_pred[i] = normal_rng(f[i], sigma);
  }
}
"

# Preparing data for stan model
stan_data <- list(
  N = length(x),
  x = x,
  y = y
)

# Compiling and fitting model
fit <- stan(model_code = stan_code, data = stan_data, iter = 2000, chains = 4)

# Printing summary
print(fit, pars = c("alpha", "lambda", "sigma"))
```

## **7.2 Compute the posterior average and variance for tree circumference at 2 years (730 days) of age.**

```{r}
# Define the point of interest
x_new <- 730  # Age in days

# Extract posterior samples
f_samples <- extract(fit, pars = "f")$f
alpha_samples <- extract(fit, pars = "alpha")$alpha
lambda_samples <- extract(fit, pars = "lambda")$lambda
sigma_samples <- extract(fit, pars = "sigma")$sigma

# Compute the posterior mean and variance of f(x_new)
N <- length(stan_data$x)
num_samples <- length(alpha_samples)
f_new_samples <- numeric(num_samples)

for (s in 1:num_samples) {
  # Squared exponential kernel for x_new
  k_star <- exp(-0.5 * (stan_data$x - x_new)^2 / lambda_samples[s]^2) * alpha_samples[s]^2
  K <- matrix(0, N, N)
  for (i in 1:N) {
    for (j in 1:N) {
      K[i, j] <- alpha_samples[s]^2 * exp(-0.5 * (stan_data$x[i] - stan_data$x[j])^2 / lambda_samples[s]^2)
    }
  }
  K <- K + diag(sigma_samples[s]^2, N)
  
  # Predictive mean and variance
  K_inv <- solve(K)
  mu_f_new <- t(k_star) %*% K_inv %*% f_samples[s, ]
  var_f_new <- alpha_samples[s]^2 - t(k_star) %*% K_inv %*% k_star
  
  # Sample from the posterior predictive distribution
  f_new_samples[s] <- rnorm(1, mean = mu_f_new, sd = sqrt(var_f_new))
}

# Transform back to circumference scale
circumference_samples <- exp(f_new_samples)

# Compute posterior average and variance
posterior_mean <- mean(circumference_samples)
posterior_variance <- var(circumference_samples)

# Display results
cat("Posterior mean circumference at 730 days:", posterior_mean, "\n")
cat("Posterior variance circumference at 730 days:", posterior_variance, "\n")

```

### **7.3 Plot the posterior distribution of exp(f). Include the data points and the posterior average computed in 7.2 in the same graph.**
```{r}
# Load required library for plotting
library(ggplot2)

# Assuming data has been loaded into `data` as per your dataset

# Compute log of circumference from the data
log_circumference <- log(data$circumference)

# Assuming posterior mean and variance have been computed as `posterior_mean` and `posterior_variance`
circumference_mean <- posterior_mean  # Computed in 7.2
circumference_variance <- posterior_variance  # Computed in 7.2
circumference_sd <- sqrt(circumference_variance)

# Create a density plot of the posterior samples
circumference_density <- density(circumference_samples)

# Plot the posterior distribution
ggplot() +
  # Add the posterior density
  geom_line(
    aes(x = circumference_density$x, y = circumference_density$y),
    color = "blue",
    linewidth = 1,   # Use 'linewidth' instead of 'size'
    alpha = 0.8
  ) +
  # Add vertical line for the posterior mean
  geom_vline(
    aes(xintercept = circumference_mean),
    color = "red",
    linetype = "dashed",
    linewidth = 1,   # Use 'linewidth' instead of 'size'
    alpha = 0.8
  ) +
  # Add data points (log-transformed circumference)
  geom_point(
    aes(
      x = exp(log_circumference),
      y = rep(0, length(log_circumference))
    ),
    color = "darkgreen",
    size = 2,
    alpha = 0.6
  ) +
  # Customize plot appearance
  labs(
    title = "Posterior Distribution of exp(f)",
    x = "Circumference (mm)",
    y = "Density",
    caption = "Posterior mean is highlighted in red."
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    axis.title = element_text(size = 14)
  )

```

