---
title: "Assignment 2 Report"
author: "Ayesha Zafar"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
---

# 2. Grid approximation: Exponential

---
Consider the following statistical model:

y∼Exp(ea+bx)

a,b∼N(0,1),

where Exp is the exponential distribution.

### **2.1 Generate data from this model. Use sample size N=50, generate values for a and b from N(0,1). Generate the x-values from Uniform(0,2).**

Below code generates the data from given model:
```{r}
# Settng seed for reproducibility
set.seed(123)

# Setting sample size
N <- 50

# Generating values for a and b from N(0, 1)
a <- rnorm(N, mean = 0, sd = 1)
b <- rnorm(N, mean = 0, sd = 1)

# Generating x-values from Uniform(0, 2)
x <- runif(N, min = 0, max = 2)

# Calculating y-values using the exponential distribution
y <- rexp(N, rate = exp(a + b * x))

# Creating data frame for easy viewing
data <- data.frame(x = x, y = y, a = a, b = b)
head(data)

```

### **2.2 Implement the grid approximation for this model.**

To implement the grid approximation for this model, we need to:

1. Create a grid for possible values of a and b.
2. Calculate the likelihood for each combination of a and b.
3. Use Bayes theorem to calculate the posterior.

```{r}
# Setting grid for parameters a and b
a_grid <- seq(-3, 3, length.out = 100)
b_grid <- seq(-3, 3, length.out = 100)

# Storing posterior in a matrix
posterior <- matrix(0, nrow = length(a_grid), ncol = length(b_grid))

# Calculating likelihood for combinations of a and b
for (i in 1:length(a_grid)) {
  for (j in 1:length(b_grid)) {
    # Rate parameter for Exp distribution
    lambda <- exp(a_grid[i] + b_grid[j] * x)
    # Likelihood of y given a, b
    likelihood <- prod(dexp(y, rate = lambda))
    
    # Calculating prior (assuming normal prior, mean 0 and std 1)
    prior_a <- dnorm(a_grid[i], mean = 0, sd = 1)
    prior_b <- dnorm(b_grid[j], mean = 0, sd = 1)
    
    # Calculating up to constant factor posterior
    posterior[i, j] <- likelihood * prior_a * prior_b
  }
}

# Normalizing posterior
posterior <- posterior / sum(posterior)

# Plotting posterior distribution
library(ggplot2)
posterior_df <- expand.grid(a = a_grid, b = b_grid)
posterior_df$posterior <- as.vector(posterior)

ggplot(posterior_df, aes(x = a, y = b, fill = posterior)) +
  geom_tile() +
  labs(title = "Posterior Distribution", x = "a", y = "b") +
  scale_fill_viridis_c() +
  theme_minimal()

```


### **2.3 Plot the full posterior distribution. What information is there in the full posterior that is lost in the marginal posteriors of a and b?**

The full posterior distribution can be plotted directly from the previous step. We’ve already generated the posterior for each combination of a and b, and visualized it using geom_tile in ggplot2.

However, the marginal posteriors of a and b would be obtained by integrating over the other parameter, which effectively collapses the full posterior into two one-dimensional distributions. By comparing the full posterior and the marginal posteriors, we can see the information that is lost.

```{r}
# Full posterior distribution
ggplot(posterior_df, aes(x = a, y = b, fill = posterior)) +
  geom_tile() +
  labs(title = "Full Posterior Distribution", x = "a", y = "b") +
  scale_fill_viridis_c() +
  theme_minimal()


# Marginal posteriors
marginal_a <- apply(posterior, 1, sum)
marginal_b <- apply(posterior, 2, sum)
par(mfrow = c(1, 2))
plot(a_grid, marginal_a, type = "l", main = "Marginal Posterior of a", xlab = "a", ylab = "Density")
plot(b_grid, marginal_b, type = "l", main = "Marginal Posterior of b", xlab = "b", ylab = "Density")
```

**Comparison Between Full and Marginal Posteriors**

| Feature                      | Full Posterior Distribution                            | Marginal Posterior Distribution                          |
|------------------------------|--------------------------------------------------------|----------------------------------------------------------|
| **Definition**                | The joint distribution of both \(a\) and \(b\).        | The individual distributions of \(a\) and \(b\), marginalizing over the other parameter. |
| **What It Shows**             | The relationship between \(a\) and \(b\), including their dependencies. | The distribution of \(a\) and \(b\) independently, ignoring their joint relationship. |
| **How It Is Visualized**      | Heatmap or 2D plot showing the joint distribution.    | Separate 1D plots showing the distribution of \(a\) and \(b\). |
| **What Information It Loses** | None – it retains the full joint relationship.         | It loses the joint relationship between \(a\) and \(b\), showing them as if they are independent. |
| **Usefulness**                | Helps in understanding how \(a\) and \(b\) interact.  | Useful for marginal inference, focusing on each parameter independently. |
| **Example Plot**              | 2D Heatmap/Contour plot for joint distribution.        | Two separate line plots for \(a\) and \(b\). |

### **2.4 Compute the marginal posterior mode of a.**

The mode of the marginal posterior of a is the value of a that maximizes the marginal posterior distribution. We can find this by identifying the maximum value in the marginal distribution of a.

```{r}
mode_a <- a_grid[which.max(marginal_a)]
cat("Mode of the marginal posterior for a:", mode_a, "\n")

# Plotting marginal posterior of a and highlighting the mode
plot(a_grid, marginal_a, type = "l", 
     main = "Marginal Posterior of a", 
     xlab = "a", ylab = "Density", 
     col = "blue", lwd = 2)

# Adding a vertical line for mode of a
abline(v = mode_a, col = "red", lwd = 2, lty = 2)

# Adding a text annotation for mode
text(mode_a, max(marginal_a), labels = paste("Mode: ", round(mode_a, 2)), 
     pos = 4, col = "red")

```


**References**

- https://rpubs.com/friendly/grid_approx

- https://en.wikipedia.org/wiki/Posterior_probability

- https://www.statlect.com/glossary/posterior-probability

- https://www.sciencedirect.com/topics/mathematics/marginal-posterior-distribution



