---
title: "Assignment 3 Report"
author: "Ayesha Zafar"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
---

# 3. Stan: E.coli

---

### **3.1 Build a Stan program to estimate the parameters K,r and the measurement error σ2. Use non-uniform priors.**

The logistic growth model describes population dynamics using the equation:

P(t) = K / [1+{(K/Po) - 1} e^−rt]

Here:

- K is the carrying capacity.
- r is the growth rate.
- Po is the initial population size (Po=0.005 is given).
- P(t) is the population size at time t.

The observed population (Pmeas) is assumed to be normally distributed around P(t) with measurement noise:

Pmeas(t)∼Normal(P(t),𝜎^2)

Using non-uniform priors for the parameters:

- K∼Normal(0.5,0.2)
- r∼Normal(0.1,0.05)
- σ∼Normal(0.1,0.05)

Below is the stan code, posterior analysis and visualization.
```{r}
library(rstan)

# Loading data from ecoli3.txt
data <- read.csv("ecoli3.txt", header = TRUE)
head(data)

# Extracting necessary variables from data
# Number of observations in the data
N <- nrow(data)          
# Time points in data
t <- data$time           
# Measured population sizes
P_meas <- data$E4        
# Initial population size
P0 <- 0.005              

# Stan model
stan_model_code <- "
data {
  // No of observations
  int<lower=0> N;               
  // Time points
  vector[N] t;                  
  // Measured population sizes
  vector[N] P_meas;             
  // Initial population size
  real<lower=0> P0;             
}
parameters {
  // Carrying capacity
  real<lower=0> K;              
  // Growth rate
  real<lower=0> r;              
  // Measurement noise
  real<lower=0> sigma;          
}
model {
  // Predicted population sizes
  vector[N] P_pred;             
  for (i in 1:N) {
    P_pred[i] = K / (1 + (K / P0 - 1) * exp(-r * t[i]));
  }
  // P_meas
  P_meas ~ normal(P_pred, sigma);  
  // Prior for K
  K ~ normal(0.5, 0.2);            
  // Prior for r
  r ~ normal(0.1, 0.05);           
  // Prior for sigma
  sigma ~ normal(0.1, 0.05);       
}
"

# Compiling stan model
stan_model <- stan_model(model_code = stan_model_code)

# Sampling stan model
fit <- sampling(
  object = stan_model,
  data = list(N = N, t = t, P_meas = P_meas, P0 = P0),
  iter = 2000,
  chains = 4,
  seed = 123
)

# Printing model results
print(fit)

```

### **3.2 Plot histogram of the marginal posterior samples of r and include the marginal posterior mean of r as a vertical line in the plot.**

```{r}
# Extracting posterior samples
posterior_samples <- extract(fit)
r_samples <- posterior_samples$r

# Calculating mean of posterior samples for r
r_mean <- mean(r_samples)

# Plotting histogram
hist(r_samples, breaks = 30, probability = TRUE, main = "Marginal Posterior of r", xlab = "r", col = "lightblue")

# Adding posterior mean as a vertical line
abline(v = r_mean, col = "red", lwd = 2, lty = 2)
legend("topright", legend = c("Posterior Mean"), col = "red", lty = 2, lwd = 2)
```

### **3.3 Determine the posterior probability that r<0.25.**

```{r}
# Calculating posterior probability where r < 0.25
posterior_prob_r <- mean(r_samples < 0.25)
cat("Posterior probability (r < 0.25):", posterior_prob_r, "\n")
```


**References**

- https://fieldnotess.medium.com/uninformative-priors-informative-priors-cef9c3c6cbdf

- https://rdrr.io/cran/nimbleCarbon/man/postHPDplot.html

