---
title: "Assignment 4 Report"
author: "Ayesha Zafar"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
---

# 4. Stan: Indomethacin

---

### **4.1 Implement a Stan program for the model. In the generated quantities block, generate the population distribution for λ.**

The below stan model estimates the decay rate λ for each subject using a hierarchical structure. The exponential decay model is used to relate concentration of indomethacin to time. Generated quantities block in stan is used to obtain population distribution of λ by saving the λ values for each subject.

```{r}
library(rstan)
library(ggplot2)
library(dplyr)

# Stan model
stan_model_code <- "
data {
  // No of observations
  int<lower=1> N;            
  // No of subjects
  int<lower=1> J;            
  // Subject index
  int<lower=1> subject[N];   
  // Time measurements
  vector[N] time;            
  // Observed concentrations
  vector[N] conc;            
}

parameters {
  // Initial concentration
  real<lower=0> C0;                       
  // Measurement noise
  real<lower=0> sigma;                    
  // Population mean for λ
  real<lower=0> lambda_mu;                
  // Population SD for λ
  real<lower=0> lambda_sigma;             
  // Subject specific decay rates
  vector<lower=0>[J] lambda;              
}

model {
  // Prior for C0
  C0 ~ normal(0, 10);                     
  // Prior for measurement noise
  sigma ~ normal(0, 1);                   
  // Hyperprior for lambda_mu
  lambda_mu ~ gamma(2, 1);                
  // Hyperprior for lambda_sigma
  lambda_sigma ~ gamma(2, 1);             
  // Subject specific priors
  lambda ~ gamma(lambda_mu, lambda_sigma); 

  conc ~ normal(C0 * exp(-lambda[subject] .* time), sigma);
}

generated quantities {
  // Posterior predictive samples
  vector[N] y_pred;                       
  // Population λ samples
  vector[1000] population_lambda;         
  for (n in 1:N) {
    y_pred[n] = normal_rng(C0 * exp(-lambda[subject[n]] * time[n]), sigma);
  }
  for (i in 1:1000) {
    population_lambda[i] = gamma_rng(lambda_mu, lambda_sigma);
  }
}
"

# Loading data from indometh.txt
indometh <- read.csv("indometh.txt")

# Extracting necessary variables from data
N <- nrow(indometh)
J <- length(unique(indometh$Subject))
subject <- as.integer(indometh$Subject)
time <- indometh$time
conc <- indometh$conc

# Preparing data for stan model
stan_data <- list(
  N = N,
  J = J,
  subject = subject,
  time = time,
  conc = conc
)

# Compiling stan model
fit <- stan(
  model_code = stan_model_code, 
  data = stan_data,
  iter = 2000,
  chains = 4
)

# Printing results
print(fit)
```

### **4.2 Plot a histogram of the population distribution samples. Include marginal posterior means of the subject-specific λi in the same figure as vertical lines.**

```{r}
# Extracting results
samples <- extract(fit)

# Plotting histogram
pop_lambda <- samples$population_lambda
marginal_means <- colMeans(samples$lambda)
ggplot() +
  geom_histogram(aes(x = pop_lambda), bins = 30, fill = "blue", alpha = 0.5) +
  geom_vline(aes(xintercept = marginal_means), color = "red", linetype = "dashed") +
  labs(
    title = "Population Distribution of Lambda",
    x = "Lambda",
    y = "Frequency"
  )

```

### **4.3 Assume a new subject enrolls in the study. Each sample from the posterior predictive distribution corresponds to a trajectory of Indomethacin concentration. Generate a plot of these trajectories.**

```{r}
# Assuming mean for new subject
new_subject_lambda <- mean(pop_lambda) 
time_seq <- seq(0, 8, by = 0.1)
posterior_predictive <- sapply(
  time_seq,
  function(t) rnorm(1000, samples$C0 * exp(-new_subject_lambda * t), mean(samples$sigma))
)

# Plotting trajectories
matplot(
  time_seq,
  t(posterior_predictive),
  type = "l",
  col = rgb(0, 0, 1, alpha = 0.2),
  lty = 1,
  xlab = "Time",
  ylab = "Concentration",
  main = "Posterior Predictive Trajectories"
)
```

### **4.4 There is a time point t∗ after which, at each t≥t∗, the predicted concentration for the new subject is less than 0.1 with >90% probability. Find t∗ to a precision of 0.01.**

```{r}
# Assuming mean for new subjects lambda
new_subject_lambda <- mean(pop_lambda)

# Setting time sequence with precision of 0.01
time_seq <- seq(0, 8, by = 0.01)

# Posterior predictive concentrations for each time point
posterior_predictive <- sapply(
  time_seq,
  function(t) rnorm(1000, samples$C0 * exp(-new_subject_lambda * t), mean(samples$sigma))
)

# Calculate probabilities for concentrations below 0.1
probabilities <- apply(posterior_predictive, 2, function(x) mean(x < 0.1))

# Finding first time point t* where P > 90% for all t >= t*
t_star_index <- which(cumprod(probabilities > 0.9) == 1)[1]

# Extracting t* if available
if (!is.na(t_star_index)) {
  t_star <- time_seq[t_star_index]
  cat("t* =", t_star, "\n")
} else {
  cat("No time point satisfies condition for all t >= t*.\n")
}
```


**References**

- https://rdrr.io/cran/nimbleCarbon/man/postHPDplot.html

- https://stackoverflow.com/questions/16571390/making-a-trajectory-plot-using-r


